Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-22 mr-review.Rmd
added ivw eq
minor updates to text

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
weighting" because each causal SNP's effect is weighted by the inverse of the variance 
of its marginal (univariate) causal effect, and the overall causal effect is the sum of the weighted SNP
causal effects.

>>>>

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
weighting" because each causal SNP's effect is weighted by the inverse of the variance 
of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
causal effects.

Specifically, @burgess_mendelian_2013 combined causal effect (ratio) estimates from $M$ independent SNPs by using Equation \@ref(eq:ivw).

\begin{equation}
  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
  (\#eq:ivw)
\end{equation}

@burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or more of the three core assumptions.

@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted profile score to accomplish statistical inference in two-sample MR analyses with GWAS data.
